Engineering momentum profiles of cold-atom beams

### 2019-02-13 Paper/version2/paper.tex
New draft...

\DeclareMathOperator*{\argmin}{argmin}

@@

{\it Procedure for Finding a Link Potential. --}
We find an appropriate link potential $V_0(x) \equiv V(x,\bm{\theta}^*)$ by performing a global search over a family of possible potentials $V(x,\bm{\theta})$ for the parameters $\bm{\theta}^*$ that reduce the $k$-integrated squared error between the desired transmission-momentum profile $T_0(k)$ and the actual profile $T_{\bm\theta}(k)$ determined by a sample potential $V(x, \bm{\theta})$. Concretely, we minimize the cost
\begin{equation}\label{eq:cost1}
  J_{\bm{\theta}} = \sum_kw_k\left|T_0(k) - T_{\bm{\theta}}(k)\right|^2,
\end{equation}
where the $k$-integral has been approximated (up to a constant factor) by a sum over a discrete set of momentum values, and $w_k$ is the weight given to momentum $k$. The weights are chosen to emphasize regions of $k$ during the minimization. For instance, when trying to create a potential barrow band-pass filter that forbids transmission for all $k$ except in the neighborhood of a chosen value $k_0$ (see Fig.~\ref{fig:Figure1} and Fig.~\ref{fig:method_illustration}, top), then it would be appropriate to increase the weights in the region of $k_0$. The convergence of our approach was somewhat sensitive to the choice of $w_k$. To arrive at acceptible solutions, we progressively increased $w_k$ in the region of $k_0$ until we found solutions that effectively balanced the need to allow flux near $k_0$ and forbid flux far from $k_0$.

With the goal of discovering experimentally viable solutions, we parameterize the family of link potentials $V(x, \bm{\theta})$, as a sum of $N$ Gaussians, each of the form
\begin{equation}\label{eq:V-param}
V_i(x; A_i, \mu_i, \sigma_i) = \frac{A_i}{\sqrt{2\pi\sigma_i^2}}\exp\left[{-\frac{(x-\mu_i)^2}{2\sigma_i^2}}\right];
\end{equation}
the parameter $\bm{\theta}$ denotes the parameter space $\{A_1, \mu_1, \sigma_1,...,A_N, \mu_N, \sigma_N\}$. While minimizing Eq.~\eqref{eq:cost1}, we enforce the parameter constraints listed in Table \ref{tab:constraints}. In addition to these constraints on the parameters, we enforce an constraint by requiring that the link potential should not extend beyond the region of potential support $x\in[-x_0,x_0]$. 
To accomplish this, we minimize the augmented cost
\begin{equation}\label{eq:cost2}
  J_{\bm{\theta}}^{\mathrm{aug}} = J_{\bm{\theta}} + \alpha \sum_i^N\int\limits_{|x|>x_0}dx\,|V_i(x; A_i,\mu_i,\sigma_i)|^2,
\end{equation}
where $\alpha$ is a tuning parameter chosen to make the added term of a similar order to $J_{\mathbf{\theta}}$ in the scenario that the potential extends outside the support region. Beyond this, we find that results are not very sensitive to the choice of $\alpha$ probably due to the very short tails of the Gaussian potentials. The integral in Eq.~\eqref{eq:cost2} evaluates to the complementary error function. The full form is given in the suppl.~material.

@@

For a choice of $\bm{\theta}$ (and hence $V(x;\bm{\theta})$), we solve for $T_{\bm{\theta}}(k)$ by integrating the Schr{\"o}dinger equation~(\ref{eq:schr}) across the region of the potential and calculating the ratio of the transmitted to the incident flux. In order to do this efficiently, we discretize the second derivative in Eq.~(\ref{eq:schr}), which transforms Eq.~(\ref{eq:schr}) into
a banded linear system of equations solvable in $O(M)$ time where $M$ is the number of $x$-steps. Using these techniques we are able to evaluate the transmission coefficient {\color{red} x} thousand times per second on a 7th generation Intel Core i7 processor under the conditions used in our tests.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{figure}
   \includegraphics[width=1\linewidth]{figures/plot_transport_profiles.pdf}
   \includegraphics[width=1\linewidth]{figures/plot_link_potential.pdf}
 \caption[Narrow band-pass filter link potential]{Two-Gaussian solution for the $k=1$ narrow band-pass filter transport profile. The top figure shows the target profile (dot-dashed, red) used during optimization and the actual profile resulting from the optimization procedure. The inset is a zoomed-in view of the region around the peak near $k=1$. The bottom figure shows the optimal link potential $V(x, \bm{\theta^*})$. {\color{blue} bold font, $k>2$?}}
 \label{fig:method_illustration}
\end{figure}

@@

 We minimize $J_{\bm{\theta}}^{\mathrm{aug}}$ for $\bm{\theta}^*$ using the global optimization routine called Differential Evolution (DE) \cite{original DE paper}. This evolutionary-based search algorithm is suitable given the non-convex (multiple local minima) nature of the optimization problem. Despite its simplicity, DE does a good job of balancing exploration of the space of link potentials against the need to efficiently learn from each sample with little tuning of the model settings. Empirically, we found DE to perform much better than several other approaches including random search, Nelder-Mead, and Simulated Annealing \cite{Tests performed using mathematica}.

{\it Narrow band-pass filter transport profile. --}
To illustrate the method described above we optimize for a narrow band-pass filter transport profile sharply-peaked near $k=k_0$. {\color{green} For simplicity, in what follows we adopt the units in which $k_0=\hbar=2m=1$}. For our target transport profile, we use a Lorentz distribution peaked at $k=1$ and with a small {\color{red} HWHM} (see Figure~\ref{fig:method_illustration}, top). For the constraints shown in Table~\ref{tab:constraints}, {\color{red} we use $\sigma_{\mathrm{min}}=0.X$, $\sigma_{\mathrm{max}}=Y.Y$, $A_{\mathrm{min}}=0.X$, and $A_{\mathrm{max}}=Y.Y$}. We simplify the optimization by searching over two-Gaussian link-potentials with equal amplitudes and widths.
{\color{green} We anticipate that this potential might be the easiest to realize in the laboratory. In addition, it is straithforward to visualize the transfer using quasi-discrete levels 
in the link potential. 
Note also that even though we work here with a very simple example that has only three unknown parameters, a method for globally searching the space of link potential is still required, because
the cost function for the optimization has many local minima corresponding to the many ways that a resonance state can be placed near the scattering energy $k^2$. }

Figure \ref{fig:method_illustration} shows the link potential and transport profile resulting for the band-pass filter optimization. The solution does a good job of suppressing transport except near $k=1$ as set by our Lorentz target profile and near $k=2$ resulting from a second resonance in the scattering potential. This additional resonance need not be a problem if the scattering atoms are sourced from a thermal reservoir with sufficiently low population near $k=2$~\footnote{In general, one needs to include the momentum distribution of the reservoir, $n(k)$,
in the cost function. To this end, one might multiply in Eq.~(\ref{eq:cost1}) $T_{\bm{\theta}}(k)$ by $n(k)$; the rest of the discussion is untouched.
The function $n(k)$ can be determined by the Bose-Einstein or Fermi-Dirac distributions. The illustration in the text practically corresponds to a Fermi gas 
at zero temperature, i.e., $n(k<k_F)=1$ and zero otherwise, with the Fermi energy that satisfies $1\ll k_F^2 \ll A_\mathrm{min}$}.
 {\color{green}  Note that the two-Gaussian solution struggles to reproduce the width of the Lorentz target profile (see the inset in Fig.~\ref{fig:method_illustration}), and the second resonance might be used instead in practical applications. Indeed, the width of the first resonance is quite narrow, 
which requires long experimental runtime; for example, a reasonable assumption $\hbar^2k_0^2/(2m)=k_B\times 100$nK, where $k_B$ is the Boltzmann constant leads to 
a single-particle time scale for tunnelling $\tau\sim 0.2 s$. In case this time scale is not experimentally realistic, the second resonance 
can be used.} Though we have not demonstrated that the solution shown in Fig.~\ref{fig:method_illustration} corresponds to a global minimum, this failure to reproduce the width may result from insufficient freedom in the family of constrained two-Gaussian potentials {\color{blue} explain better?}.

@@

Though we have demonstrated our optimization method in a very simple scenario, it is possible to apply this technique to more complicated scenarios such as double band-pass filter or step transport profiles for various experimental scenarios. We find that these more complicated transport profiles require more than two Gaussian potentials and that the solutions using the procedure we describe are highly-sensitive to the potential parameters. We leave the thorough exploration of these ideas to future work. In the next section, we discuss a possible experimental application of the narrow band-pass filter solution.

@@

\subsection{Cost Function}\label{sec:cost_function}

\begin{equation}
  J'_{\mathbf{\theta}} = J_{\mathbf{\theta}} + \alpha J_{\mathrm{boundary}},
\end{equation}
where $\alpha>0$ is the weight given to cost incurred for having the potential extend outside the support region of the potential.

>>>>

\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator\erf{erf}
\DeclareMathOperator\erfc{erfc}

@@

{\it Procedure for Finding a Link Potential. --}
We find an appropriate link potential $V_0(x) \equiv V(x,\bm{\theta}^*)$ by performing a global search over a family of possible potentials $V(x,\bm{\theta})$ for the parameters $\bm{\theta}^*$ that reduce the $k$-integrated squared error between the desired transmission-momentum profile $T_0(k)$ and the actual profile $T_{\bm\theta}(k)$ determined by a sample potential $V(x, \bm{\theta})$. Concretely, we minimize the cost
\begin{equation}\label{eq:cost1}
  J_{\bm{\theta}} = \sum_kw_k\left|T_0(k) - T_{\bm{\theta}}(k)\right|^2,
\end{equation}
where the $k$-integral has been approximated (up to a constant factor) by a sum over a discrete set of momentum values, and $w_k$ is the weight given to momentum $k$. The weights are chosen to emphasize regions of $k$ during the minimization. For instance, when trying to create a potential barrow band-pass filter that forbids transmission for all $k$ except in the neighborhood of a chosen value $k_0$ (see Fig.~\ref{fig:Figure1} and Fig.~\ref{fig:method_illustration}, top), then it would be appropriate to increase the weights in the region of $k_0$. The convergence of our approach was somewhat sensitive to the choice of $w_k$.

We developed a strategy for choosing the weights that takes into account the following considerations: 1) the transport coefficient is constrained to be zero for $k=0$ so the weight can be smaller for small $k$, 2) the transport coefficient approaches 1 for large $k$ so higher weights are required in the large-$k$ region if one wishes to supress flux at large $k$, and 3) to reproduce narrow features in the target transport profile, it may be helpful to increase the weight in the $k$-region of these features. Follwing these principles, we arrive at the following formula for the weight function:
\begin{equation}\label{eq:weight-of-k}
  w(k;r) = w_{\mathrm{bg}}(k) + rT_0(k)
\end{equation}
where $w_{\mathrm{bg}}(k)$ is chosen to account for considerations 1 and 2 above, and the term proportional to the positive constant $r$ accounts for consideration 3. For $w_{\mathrm{bg}}(k)$, we use the {\color{blue} transport profile resulting from a morse potential WITH WHAT PARAMS?}:
\begin{equation}\label{eq:weight-of-k}
  w_{\mathrm{bg}}(k) = \left[\frac{\sinh^2(\sqrt{2}\pi k)}{\sinh^2(\sqrt{2}\pi k) + \cosh^2(\sqrt{7}\pi/2)}\right]^{1/2}.
\end{equation}
The parameter $r$ was chosen by trial and error for each target profile $T_0(k)$.

With the goal of discovering experimentally viable solutions, we parameterize the family of link potentials $V(x, \bm{\theta})$, as a sum of $N$ Gaussians, each of the form
\begin{equation}\label{eq:V-param}
V_i(x; A_i, \mu_i, \sigma_i) = \frac{A_i}{\sqrt{2\pi\sigma_i^2}}\exp\left[{-\frac{(x-\mu_i)^2}{2\sigma_i^2}}\right];
\end{equation}
the parameter $\bm{\theta}$ denotes the parameter space $\{A_1, \mu_1, \sigma_1,...,A_N, \mu_N, \sigma_N\}$. While minimizing Eq.~\eqref{eq:cost1}, we enforce the parameter constraints listed in Table \ref{tab:constraints}. In addition to these constraints on the parameters, we enforce an constraint by requiring that the link potential should not extend beyond the region of potential support $x\in[-x_0,x_0]$. 
To accomplish this, we minimize the boundary-augmented cost function
\begin{equation}\label{eq:cost2}
  J_{\bm{\theta}}^{\mathrm{aug}} = J_{\bm{\theta}} + \alpha \sum_i^N\int\limits_{|x|>x_0}dx\,|V_i(x; A_i,\mu_i,\sigma_i)|^2,
\end{equation}
where $\alpha$ is a tuning parameter chosen to make the added term of a similar order to $J_{\mathbf{\theta}}$ in the scenario that the potential extends outside the support region. Beyond this, we find that results are not very sensitive to the choice of $\alpha$ probably due to the very short tails of the Gaussian potentials. The integral in Eq.~\eqref{eq:cost2} evaluates to the complementary error function. The full form of $J_{\bm{\theta}}^{\mathrm{aug}}$ is given in the suppl.~material.

@@

For a choice of $\bm{\theta}$ (and hence $V(x;\bm{\theta})$), we solve for $T_{\bm{\theta}}(k)$ by integrating the Schr{\"o}dinger equation~(\ref{eq:schr}) across the region of the potential and calculating the ratio of the transmitted to the incident flux. In order to do this efficiently, we discretize the second derivative in Eq.~(\ref{eq:schr}), which transforms Eq.~(\ref{eq:schr}) into
a banded linear system of equations solvable in $O(M)$ time where $M$ is the number of $x$-steps. Using these techniques we are able to evaluate the transmission coefficient 3.7 thousand times per second on a 7th generation Intel Core i7 processor for a test involving 1,000 randomly generated two-Gaussian potentials and 100 different scattering momenta.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% 
\begin{figure}
   \includegraphics[width=1\linewidth]{figures/plot_transport_profiles_with_inset.pdf}
 \caption[Narrow band-pass filter link potential]{Two-Gaussian solution for the $k=1$ narrow band-pass filter transport profile. The main figure shows the target profile (dot-dashed, red) used during optimization and the actual transport profile resulting from the optimization procedure (solid, blue). The inset figure shows the optimal link potential $V(x, \bm{\theta^*})$. {\color{blue} bold font, $k>2$?}}
 \label{fig:method_illustration}
\end{figure}

@@

We minimize $J_{\bm{\theta}}^{\mathrm{aug}}$ for $\bm{\theta}^*$ using the global optimization routine called Differential Evolution (DE) \cite{storn1997differential}. This evolutionary-based search algorithm is suitable given the non-convex (multiple local minima) nature of the optimization problem. Despite its simplicity, DE does a good job of balancing exploration of the space of link potentials against the need to efficiently learn from each sample with little tuning of the model settings. Empirically, we found DE to perform much better than several other approaches including random search, Nelder-Mead, and Simulated Annealing for this particular optimization problem.

{\it Narrow band-pass filter transport profile. --}
To illustrate the method described above we optimize for a narrow band-pass filter transport profile sharply-peaked near $k=k_0$. For simplicity, we adopt the units in which $k_0=\hbar=2m=1$. For our target transport profile, we use a Lorentz profile
\begin{equation}\label{eq:Ttarget}
  T(k; k_0,b) = \left[1 + \frac{(k-k_0)^2}{b^2}\right]^{-1}
\end{equation}
with $k_0=1$ and $b=0.03$ (see Figure~\ref{fig:method_illustration}). For the constraints shown in Table~\ref{tab:constraints}, we use $\sigma_{\mathrm{min}}=0.2$, $\sigma_{\mathrm{max}}=3$, $A_{\mathrm{min}}=5$, and $A_{\mathrm{max}}=30$. We further simplify the optimization by searching over two-Gaussian link-potentials with equal amplitudes and widths. We anticipate that this potential might be the easiest to realize in the laboratory. In addition, it is straithforward to visualize the transfer using quasi-discrete levels in the link potential. Note also that even though we work here with a very simple example that has only three unknown parameters, a method for globally searching the space of link potential is still required, because the cost function for the optimization has many local minima corresponding to the many ways that a resonance state can be placed near the scattering energy $k^2$.

Figure \ref{fig:method_illustration} shows the link potential and transport profile resulting for the band-pass filter optimization. The solution does a good job of suppressing transport except near $k=1$ as set by our Lorentz target profile and near $k=2$ resulting from a second resonance in the scattering potential. This additional resonance need not be a problem if the scattering atoms are sourced from a thermal reservoir with sufficiently low population near $k=2$~\footnote{In general, one needs to include the momentum distribution of the reservoir, $n(k)$,
in the cost function. To this end, one might multiply in Eq.~(\ref{eq:cost1}) $T_{\bm{\theta}}(k)$ by $n(k)$; the rest of the discussion is untouched.
The function $n(k)$ can be determined by the Bose-Einstein or Fermi-Dirac distributions. The illustration in the text practically corresponds to a Fermi gas 
at zero temperature, i.e., $n(k<k_F)=1$ and zero otherwise, with the Fermi energy that satisfies $1\ll k_F^2 \ll A_\mathrm{min}$}.
 {\color{red}  Note that the two-Gaussian solution struggles to reproduce the width of the Lorentz target profile (see the inset in Fig.~\ref{fig:method_illustration}), and the second resonance might be used instead in practical applications. Indeed, the width of the first resonance is quite narrow, 
which requires long experimental runtime; for example, a reasonable assumption $\hbar^2k_0^2/(2m)=k_B\times 100$nK, where $k_B$ is the Boltzmann constant leads to 
a single-particle time scale for tunnelling $\tau\sim 0.2 s$. In case this time scale is not experimentally realistic, the second resonance 
can be used. Though we have not demonstrated that the solution shown in Fig.~\ref{fig:method_illustration} corresponds to a global minimum, this failure to reproduce the width may result from insufficient freedom in the family of constrained two-Gaussian potentials. }

@@

Though we have demonstrated our optimization method in a very simple scenario, it is possible to apply this technique to more complicated scenarios such as a double band-pass filter or step transport profiles for various experimental scenarios. We find that these more complicated transport profiles require more than two Gaussian potentials and that the solutions using the procedure we describe are highly-sensitive to the potential parameters. We leave the thorough exploration of these ideas to future work. In the next section, we discuss a possible experimental application of the narrow band-pass filter solution.

@@

\subsection{Boundary-Augmented Cost Function}\label{sec:Jaug}
Equation \eqref{eq:cost2} shows the boundary-augmented cost function which includes a term that increases the cost for link-potential solutions that extend beyond the support region $x\in[-x_0,x_0]$. This added terms is
\begin{equation}
  J_{\mathrm{boundary}} = \alpha \sum_i^N J_{\mathrm{boundary}}^i,
\end{equation}
where 
\begin{equation}\label{eq:Jboundaryi}
  J_{\mathrm{boundary}}^i = \int\limits_{|x|>x_0}dx\,|V_i(x; A_i,\mu_i,\sigma_i)|^2.
\end{equation}
Assuming the Gaussian potential form in Eq.~\eqref{eq:V-param}, this evaluates to
\begin{equation}\label{eq:JboundaryIGaussian}
  J_{\mathrm{boundary}}^i = \frac{\sqrt{\pi}}{2}A_i^2\sigma_i\left[
    \erfc\left(\frac{x_0+\mu_i}{\sigma_i}\right) +
    \erfc\left(\frac{x_0-\mu_i}{\sigma_i}\right)
  \right],
\end{equation}
where $\erfc$ is the complementary error function.
